---
title: "Gender Neutral Names"
author: "Vishal Lala"
date: "1/17/2022"
output:
  html_document: 
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cerulean
    highlight: pygments
    df_print: paged
  prettydoc::html_pretty:
    theme: cayman
    toc: true
    df_print: paged
  tufte::tufte_handout: default
  ioslides_presentation: default
  pdf_document:
    toc: true
    toc_depth: 3
    highlight: pygments
  html_vignette: default
  beamer_presentation: default
---

Over the last couple of decades, the line distinguishing gender roles has blurred. Increasingly, men are taking on tasks and jobs that were previously linked to women and a growing number of women are happy to let the husband manage the kids while she becomes the primary bread-getter for the family. Men are no longer shying from jobs in nursing and women are taking on manufacturing jobs. 

Has the evolution of our thinking of gender stereotypes had an impact on how parents name their kids? Over the last decade, names such as Charlie, Sklyer, and Dakota have been given to boys and girls in approximately equal proportion. The question we examine is whether there has been an increase in such gender-neutral names over time. We define gender-neutral as being a name for one gender is also given to at least 25% of the other gender. For instance, a name given to 30% male babies (and therefore 70% female babies) would qualify as a gender-neutral name. 

The data for this study was obtained from the [Social Security Administration's (SSA) website](https://www.ssa.gov/oact/babynames/) which has shared data on names of babies and their gender going back more than a 100 years. The yearly data shared by SSA was first combined into a single file.  
The data was gathered from the SSA on `r library(lubridate); dt <- ymd(Sys.Date()); paste0(month(dt,label = T),' ',day(dt),', ',year(dt))`



```{r warning=F, message=F}
data_us = read.csv('https://raw.githubusercontent.com/pseudorational/datasets/main/baby_names.csv')
library(dplyr); library(ggplot2)
dat_us = 
  inner_join(data_us[data_us$gender=='M',],
           data_us[data_us$gender=='F',],
           by=c('name'='name','year'='year'))%>%
  mutate(pct_male = 100*(n.x/(n.x+n.y)))%>%
  filter(pct_male>=25 & pct_male <=75)%>%
  group_by(year)%>%
  count()%>%
  ungroup()%>%
  inner_join(
    data_us %>%group_by(year)%>%summarize(total_n = n())%>%ungroup(), by = c('year'='year'))%>%
  mutate(pct_gender_neutral = (n/total_n)*100)%>%
  select(year, pct_gender_neutral)%>%
  right_join(data.frame(year = 1880:2020,pct_gender_neutral = 0),by = c('year'='year'))%>%
  mutate(pct_gender_neutral = ifelse(is.na(pct_gender_neutral.x), pct_gender_neutral.y, pct_gender_neutral.x))%>%
  arrange(year)%>%
  select(year, pct_gender_neutral)
dat_us


dat_us%>%
  ggplot(aes(x=year, y = pct_gender_neutral))+
  geom_point()+
  ggthemes::theme_wsj()
```


Here are the Top 5 Gender-Neutral Names by Year. Note, only names that qualify our definition of gender-neutral are included, so for certain years, there may be fewer than 5 names. 
```{r warning=F, message=F}
inner_join(data_us[data_us$gender=='M',],
           data_us[data_us$gender=='F',],
           by=c('name'='name','year'='year'))%>%
  mutate(pct_male = 100*(n.x/(n.x+n.y)))%>%
  filter(pct_male>=25 & pct_male <=75)%>%
  filter(n.x>1000)%>%
  mutate(gen_neutral = abs(pct_male - 50))%>%
  group_by(year)%>%
  slice_min(gen_neutral,n = 4)%>%
  ungroup()%>%
  arrange(desc(year))%>%
  select(year, name)

  
```

Next, we explore the trends in gender-neutral names across different States. 
```{r warning=F, message=F}
library(dplyr); library(ggplot2)
data = read.csv('C:/myFiles/Learning/datasets/baby_names_by_state.csv')

dat = 
  inner_join(data[data$gender=='M',],
           data[data$gender=='F',],
           by=c('name'='name', 'year'='year','state'='state'))%>%
  mutate(pct_male = 100*(n.x/(n.x+n.y)))%>%
  filter(pct_male>=20 & pct_male <=80)%>%
  group_by(state, year)%>%
  count()%>%
  ungroup()%>%
  inner_join(
    data %>%group_by(state, year)%>%summarize(total_n = n())%>%ungroup(), by = c('year'='year', 'state'='state'))%>%
  mutate(pct_gender_neutral = (n/total_n)*100)

library(ggplot2)
library(gridExtra)
dat%>%
#  filter(state %in% c('NJ','CA','AK'))%>%
  ggplot(aes(x=year, y = pct_gender_neutral))+
#  geom_point()+
  geom_line()+
  facet_wrap(~state, scales = 'free_y')+
  ylab(label = '')+
  theme(axis.title.x=element_blank(),
        axis.text.x = element_blank(),
        axis.title.y=element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y=element_blank())

```

